Cultural Transmission and the Social Contexts of Technological Transitions during the Late Paleolithic of Korea

### 2023-05-11 analysis/paper/paper.qmd
edits to conclusion, references and methods

Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined variations in shape and the relationship between each attribute using morphological attributes on stemmed points (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yield easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness because these measurements were not available to us. For the landmark analysis, we placed 11 landmarks on the outline of each stemmed point and calculated distances between landmark coordinate pairs to derive attributes for statistical analysis. The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic. Using the point tool in ImageJ [@schneider2012nih], we captured the landmarks from images of the artifacts, and exported them as XY coordinate data for further analysis.

@@

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka. It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. We also found a slight decrease in CV values for the four attributes. Those results can be interpreted as a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skill transmission. In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques. We did not find strong support for distinctive, exclusive cultural transmission biases in tool-making in different assemblages. Correlation coefficient values across the assemblages confirm the overall chronological trend indicating indirect bias in Phase 3. Raw material choice appears to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.

One possibility is that these more easily manageable raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity. Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did]. Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf. @fitzhugh2001risk]. Another possibility is that the cooler climate might have resulted in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources <!-- citation needed, or delete this sentence --> , which would have concentrated them in fewer predictable locations. In either case, these conditions likely led to larger networks of individuals interacting with one another and an increased opportunities to learn from experienced teachers or successful models. This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.

A key difference between our work and Bettinger and Eerkens and Garvey is the time span that we are attempting to infer cultural transmission for. For Bettinger and Eerkens and Garvey the time span is in the order of 100s of years or dozens of human generations, while our study spans 1000s of years and 100s of human generations.  In her simulation study, Garvey was able to derive the population at a particular site based on the number room block (house) and assumed four learning generation for about 100 years of site occupation [@garvey2018current].

Although social learning and migration or drift models requires approximate estimates for population and generation turnover required to transmit knowledge of new technology, we were unable to employ this approach due to the large number of assumptions required to cover a 25,000-year time span. but our research sites lacked sufficient information to utilize this method. In our future study, we will further investigate how to quantify multiple variables and conduct simulations to develop a logical bridge between our primary approach borrowed from Bettinger and Eerkens [-@bettinger1999point] and social contexts for the appearance of stemmed points. We acknowledge that we did not account for the possibility of the shape of stone artifacts being modified through reuse and resharpening, which can result in the remaining shape of stemmed points being different from their original form through inherited technology (i.e. the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional]. It is our assumption that the majority of stemmed points used in our research have retained their original shape, given that retouch is rare on the tip of stemmed points. In our study, we only considered two types of biases as social contexts without considering other factors that impact on morphology of stemmed points such as tool efficiency and manufacturing economization. Mainly due to limited empirical evidence to explore a multiplicity of conjectural scenarios, we decided to focus on our primary goal, understanding the general trajectory of technological transition. In future research, we plan to focus more closely on individual assemblages and site occupation to gain a better understanding of the population responsible for creating and using stemmed points. By so doing, we anticipate achieving a more exhaustive understanding of the morphology of stemmed points and the mechanisms behind their transmission.

>>>>

Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined variations in shape and the relationship between each attribute using morphological attributes on stemmed points (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yield easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness because these measurements were not available to us. For the landmark analysis, we placed 11 landmarks on the outline of each stemmed point and calculated distances between landmark coordinate pairs to derive attributes for statistical analysis. The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic. Using the point tool in ImageJ [@schneider2012nih], we captured the landmarks from images of the artifacts, and exported them as XY coordinate data for further analysis. We consider that the stone artifacts were generally modified very little during their use life by reuse and resharpening. While retouching can alter artefact shapes independent of transmission biases (i.e. the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional], retouch is very rare on the margins and tip of stemmed points, so we consider that the majority of stemmed points used in our sample have retained their original shape.

@@

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka. This is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. We also found a slight decrease in CV values. Those results may indicate a shift away from trial and error, towards a greater reliance on copying a model in the social context of tool-making skill transmission. Raw material choice appears to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.

One possibility is that these more easily manageable raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity. Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did]. Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf. @fitzhugh2001risk]. Another possibility is that the cooler climate might have resulted in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources, which would have concentrated them in fewer predictable locations. In our previous research, we observed that forager groups with stemmed points were able to occupy higher elevation, where the temperature was generally lower [@park2022did]. The changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of increased social networks. This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.

A key difference between our work and that of Bettinger and Eerkens and Garvey, which motivated our approach, is the time span that we are attempting to infer cultural transmission for. For Bettinger and Eerkens and Garvey the time span is in the order of 100s of years or dozens of human generations, while our study spans 1000s of years and 100s of human generations. In her simulation study, Garvey was able to derive the population at a particular site based on the number room block (house) and assumed four learning generation for about 100 years of site occupation [@garvey2018current]. Our findings rely on the untested assumption that similar regularities about cultural transmission are relevant over much longer time periods. Are transmission biases scale invariant with respect to time? Or is there a time span that when exceeded, patterns in cultural transmission become so overwhelmed with noise such that they are no longer discernible? We recognize that the answer to this question is important for determining the validity of our results. @Premo_Kuhn_2010 used an agent-based model to investigate Paleolithic cultural change and found that the visibility of cultural change is strongly affected by local extinction rates, in some cases mimicking the results of conformist cultural transmission. Longer time periods may increase the probability of extinction events altering the archaeological record in this way. Hopefully our claims will stimulate future work on this question to assess the reliability of our assumptions. A good entry point to this next phase of work could be adapting Garvey's social learning models to cover a 20,000 year span, such as reported in our study. This would require drawing on archaeological and ethnographic information for approximate estimates for the population and generation turnover values required to transmit knowledge of new technology. Approximate Bayesian Computation is likely to be a productive computational method for refining these estimates, give how effective it has been in inferring models of cultural transmission from the archaeological record [@Crema_Edinborough_Kerig_Shennan_2014].
